plyranges: a grammar for transforming genomics data

### 2018-05-16 paper.Rmd
- spelling fixes - making examples flow better - consistently use \texttt for package names

\begin{figure}
\includegraphics[width=\textwidth]{diagrams/GRanges}
\caption{An illustration of the GRanges data model for a
sample from an RNA-seq experiment. The core components of the data model
inlcude a seqname column (representing the chromosome), a ranges column
which consists of start and end coordinates for a genomic region, and a
strand identifier (either positive, negative, or unstranded). Metadata
are included as columns to the right of the dotted line as annotations
(gene\_id) or range level covariates (score).}
\label{fig:GRanges} 
\end{figure}

@@

As with other genomic grammars, we define set operations that treat
ranges as sets of integers, including _intersect_, _union_,
_difference_, and _complement_. There are two sets of these: parallel
and merging. The parallel operations map to infix operators, which we
surround with _%_ symbols, a convention borrowed from R syntax.  For
example, the parallel intersection (_x %intersect% y_) finds the
intersecting range between _xi_ and _yi_ for _i_ in _1...n_, where _n_
is the length of both _x_ and _y_. In constrast, the merging
intersection (_intersect\_ranges(x, y)_) returns a new set of disjoint
ranges representing whereever there was overlap between a range in _x_
and a range in _y_. We use the infix syntax for the parallel
operations, since it is the conventional syntax for parallel, binary
operations in R. Finding the parallel union will fail when two ranges
have a gap, so we introduce a _span_ operator that takes the union
while filling any gap. The _complement_ operation is unique in that it
is unary. It finds the regions not covered by any of the ranges in a
single set. Closely related is the _between_ parallel operation, which
finds the gap separating _xi_ and _yi_. The binary operations are
callable from within arithmetic, restriction and aggregation
expressions.

@@

The design of plyranges adheres to well understood principles of
language and API design: cognitive consistency, cohesion,
expressiveness and endomorphism \cite{Green1996-qg}. 
To varying degrees, these principles also underly the
design of dplyr and the Bioconductor infrastructure.

@@

We aim for our interfaces to have a simple and direct mapping to the
user's cognitive model, i.e., how the user thinks about the data. This
requires careful selection of the level of abstraction so that the
user can express workflows in the language of genomics. This motivates
the adoption of the tidy GRanges object as our central data
structure. The basic data.frame and dplyr tibble lack any notion of
genomic ranges and so could not easily support our genomic grammar,
with its specific verbs for range-oriented data manipulation. Another
example of cognitive consistency is how plyranges is insensitive to
direction/strand by default when, e.g., detecting
overlaps. GenomicRanges has the opposite behavior. We believe that
defaulting to purely spatial overlap is most intuitive to most users.

<!--
ML:
Hadley actually presents dplyr as a 'cognitive tool', i.e., something
that helps users think about the problem. In my opinion, the best we
can do is attempt to match the intuition of the user, so when the user
decides what to do, there is an obvious way of doing it. I'm not ready
to claim that the user should start from our grammar.
-->
Like dplyr, plyranges verbs are functional: they are free of side
effects and return their result. This enables chaining of verbs
through syntax like the forward pipe operator (`%>%`, read as "then")
of the magrittr package \cite{R-magrittr}. This has syntax has a direct
cognitive mapping to natural language and the intuitive notion of
pipelines. The low-level object-oriented APIs of Bioconductor tend to
manipulate data via subreplacement functions, like `start(gr) <-
x`. These ultimately produce the side effect of replacing a symbol
mapping in the current environment and thus are not amenable to
so-called fluent syntax.

@@

A function is cohesive if it performs a singular task without
producing any side-effects. Singular tasks are not necessarily atomic;
they can always be broken down further at lower levels of abstraction.
For example, to resize a range, the user needs to specify which
position (start, end, midpoint) should be invariant over the
transformation. The `resize()` function from the GenomicRanges package
has a `fix` argument that sets the anchor, so calling `resize()`
coalesces anchoring and width modification. The coupling at the
function call level is justified since the effect of setting the width
depends on the anchor. However, plyranges increases cohesion and
decouples the anchoring into its own function call.

@@

Expressiveness relates to the information content in code: the
programmer should be able to clarify intent without unnecessary
verbosity. For example, our overlap-based join operations are more
concise than the multiple steps necessary to achieve the same effect
in the original GenomicRanges API. In other cases, the plyranges API
increases verbosity for the sake of clarity and cohesion. Explicitly
calling `anchor()` can require more typing, but the code is easier to
comprehend. Another example is the set of routines for importing
genomic annotations, including `read\_gff()`, `read\_bed()`, and
`read_bam()`. Compared to the generic `import()` in rtracklayer, the
explicit format-based naming in plyranges clarifies intent and the
type of data being returned. Similarly, every plyranges function that
computes with strand information indicates its intentions by including
suffixes such as `directed`, `upstream` or `downstream` in its name,
otherwise strand is ignored. The GenomicRanges API does not make this
distinction explicit in its function naming, instead relying on a
parameter that defaults to strand sensitivity, an arguably unintuitive
behavior.

@@

The Bioconductor package `AnnotationHub` \cite{R-ahub}  can be used to search for 
BigWig files from ChIP-Seq experiments from the Human Epigenome Roadmap project 
\cite{Roadmap-Epigenomics-Consortium2015-pr}. Here we focus on assays for primary
T CD8+ memory cells from peripheral blood. Using `plyranges` we will read the 
BigWig file  corresponding to the H3 lysine 27 trimethylation (H3K27Me3) 
methylation mark over chromosome 10.

@@

Then we read the BigWig file only extracting scores if they overlap chromosome
10. We also add the genome build information to the resulting ranges.

@@

After filtering for regions with a coverage score greater than 8, we can
reduce individual runs to ranges representing the islands of coverage,
using the `reduce_ranges()` function. This function allows a summary
to be computer over each island: in this case we take the maximum of 
the scores to find the coverage peaks over chromosome 10.

@@

We can directly obtain the genome information from the header of the
BAM file: in this case the reads were aligned to the hg19 genome build
and there are no reads overlapping the mitochondrial genome. To generate 
bins of fixed width 10000bp we apply `tile()` function from `GenomicRanges`
to the genomic coordinates (with `unlist` to return a GRanges object)
```{r bins}
locations <- h1_bam_sorted %>%
  read_bam() %>% 
  get_genome_info()

bins <- locations %>% 
  tile(width = 10000L) %>%
  unlist(use.names = FALSE)
```

Next we only read in alginments that overlap the genomic locations
we are interested in and select the query seqeunce. Note that
the reading of the BAM file is deferred: only alignments that pass the filter 
are loaded into memory. We can add another column representing the GC proportion 
for each alignment using the `letterFrequency()` function from the `Biostrings` 
package \cite{R-biostrings}.

@@

Finally, we use `group_by_overlaps()` with `summarize()` to compute the
total number of reads and averge GC content within each genomic window.

@@

We can then compute the per-chromosome transition-transversion ratio using the
`group_by` and `summarize` pattern (figure \ref{fig:titv-viz}).

@@

We have shown how to create expressive and 
reproducible genomic workflows using the `plyranges` DSL. By
realising that the GRanges data model is tidy we have highlighted how to 
implement a grammar for performing genomic arithemtic, aggregation, restriction 
and merging. Our examples show that `plyranges` code is succinct, human
readable and can take advantage of the interoperability provided by
the Bioconductor ecosytem and the R language.

>>>>

\begin{figure}
\includegraphics[width=\textwidth]{diagrams/GRanges}
\caption{An illustration of the GRanges data model for a
sample from an RNA-seq experiment. The core components of the data model
include a seqname column (representing the chromosome), a ranges column
which consists of start and end coordinates for a genomic region, and a
strand identifier (either positive, negative, or unstranded). Metadata
are included as columns to the right of the dotted line as annotations
(gene\_id) or range level covariates (score).}
\label{fig:GRanges} 
\end{figure}

@@

As with other genomic grammars, we define set operations that treat
ranges as sets of integers, including _intersect_, _union_,
_difference_, and _complement_. There are two sets of these: parallel
and merging. The parallel operations map to infix operators, which we
surround with _%_ symbols, a convention borrowed from R syntax.  For
example, the parallel intersection (_x %intersect% y_) finds the
intersecting range between _xi_ and _yi_ for _i_ in _1...n_, where _n_
is the length of both _x_ and _y_. In contrast, the merging
intersection (_intersect\_ranges(x, y)_) returns a new set of disjoint
ranges representing wherever there was overlap between a range in _x_
and a range in _y_. We use the infix syntax for the parallel
operations, since it is the conventional syntax for parallel, binary
operations in R. Finding the parallel union will fail when two ranges
have a gap, so we introduce a _span_ operator that takes the union
while filling any gap. The _complement_ operation is unique in that it
is unary. It finds the regions not covered by any of the ranges in a
single set. Closely related is the _between_ parallel operation, which
finds the gap separating _xi_ and _yi_. The binary operations are
callable from within arithmetic, restriction and aggregation
expressions.

@@

The design of `plyranges` adheres to well understood principles of
language and API design: cognitive consistency, cohesion,
expressiveness and endomorphism \cite{Green1996-qg}. 
To varying degrees, these principles also underlie the
design of `dplyr` and the Bioconductor infrastructure.

@@

We aim for our interfaces to have a simple and direct mapping to the
user's cognitive model, i.e., how the user thinks about the data. This
requires careful selection of the level of abstraction so that the
user can express workflows in the language of genomics. This motivates
the adoption of the tidy GRanges object as our central data
structure. The basic data.frame and `dplyr` tibble lack any notion of
genomic ranges and so could not easily support our genomic grammar,
with its specific verbs for range-oriented data manipulation. Another
example of cognitive consistency is how `plyranges` is insensitive to
direction/strand by default when, e.g., detecting
overlaps. `GenomicRanges` has the opposite behavior. We believe that
defaulting to purely spatial overlap is most intuitive to most users.

<!--
ML:
Hadley actually presents dplyr as a 'cognitive tool', i.e., something
that helps users think about the problem. In my opinion, the best we
can do is attempt to match the intuition of the user, so when the user
decides what to do, there is an obvious way of doing it. I'm not ready
to claim that the user should start from our grammar.
-->
Like `dplyr`, `plyranges` verbs are functional: they are free of side
effects and return their result. This enables chaining of verbs
through syntax like the forward pipe operator (`%>%`, read as "then")
of the `magrittr` package \cite{R-magrittr}. This has syntax has a direct
cognitive mapping to natural language and the intuitive notion of
pipelines. The low-level object-oriented APIs of Bioconductor tend to
manipulate data via sub-replacement functions, like `start(gr) <-
x`. These ultimately produce the side effect of replacing a symbol
mapping in the current environment and thus are not amenable to
so-called fluent syntax.

@@

A function is cohesive if it performs a singular task without
producing any side-effects. Singular tasks are not necessarily atomic;
they can always be broken down further at lower levels of abstraction.
For example, to resize a range, the user needs to specify which
position (start, end, midpoint) should be invariant over the
transformation. The `resize()` function from the `GenomicRanges` package
has a `fix` argument that sets the anchor, so calling `resize()`
coalesces anchoring and width modification. The coupling at the
function call level is justified since the effect of setting the width
depends on the anchor. However, plyranges increases cohesion and
decouples the anchoring into its own function call.

@@

Expressiveness relates to the information content in code: the
programmer should be able to clarify intent without unnecessary
verbosity. For example, our overlap-based join operations are more
concise than the multiple steps necessary to achieve the same effect
in the original `GenomicRanges` API. In other cases, the `plyranges` API
increases verbosity for the sake of clarity and cohesion. Explicitly
calling `anchor()` can require more typing, but the code is easier to
comprehend. Another example is the set of routines for importing
genomic annotations, including `read\_gff()`, `read\_bed()`, and
`read_bam()`. Compared to the generic `import()` in `rtracklayer`, the
explicit format-based naming in plyranges clarifies intent and the
type of data being returned. Similarly, every plyranges function that
computes with strand information indicates its intentions by including
suffixes such as `directed`, `upstream` or `downstream` in its name,
otherwise strand is ignored. The `GenomicRanges` API does not make this
distinction explicit in its function naming, instead relying on a
parameter that defaults to strand sensitivity, an arguably unintuitive
behavior.

@@

In the workflow of ChIP-seq data analysis, we are interested in finding peaks 
from islands of coverage over chromosome. Here we will use `plyranges`
to call peaks from islands of coverage above 8 then plot the region
surrounding the tallest peak.

Using `plyranges` and the the Bioconductor package `AnnotationHub` \cite{R-ahub}
we can download and read BigWig files from ChIP-Seq experiments from 
the Human Epigenome Roadmap project \cite{Roadmap-Epigenomics-Consortium2015-pr}. 
Here we analyse a BigWig file corresponding to H3 lysine 27 trimethylation 
(H3K27Me3) of primary T CD8+ memory cells from peripheral blood, focussing
on coverage islands over chromosome 10.

@@

Then we read the BigWig file only extracting scores if they overlap chromosome
10. We also add the genome build information to the resulting ranges. This
book-keeping is good practice as it ensures the integrity of any 
downstream operations such as finding overlaps.

@@

After filtering for regions with a coverage score greater than 8, we can
reduce individual runs to ranges representing the islands of coverage by
using the `reduce_ranges()` function. This function allows a summary
to be computer over each island: in this case we take the maximum of 
the scores to find the coverage peaks over chromosome 10.

@@

We can directly obtain the genome information from the header of the
BAM file: in this case the reads were aligned to the hg19 genome build
and there are no reads overlapping the mitochondrial genome. To generate 
bins of fixed width 10000nt we apply the `tile_ranges()` function
to the genomic coordinates in locations.
```{r bins}
locations <- h1_bam_sorted %>%
  read_bam() %>% 
  get_genome_info()

bins <- locations %>% 
  tile_ranges(width = 10000L)
```

Next we only read in alignments that overlap the genomic locations
we are interested in and select the query sequence. Note that
the reading of the BAM file is deferred: only alignments that pass the filter 
are loaded into memory. We can add another column representing the GC proportion 
for each alignment using the `letterFrequency()` function from the `Biostrings` 
package \cite{R-biostrings}.

@@

Finally, we use `group_by_overlaps()` to see where the alignments overlap the 
genomic windows and then apply `summarize()` to compute the
total number of reads and average GC content within each window.

@@

We then compute the transition-transversion ratio over each chromosome using
`group_by()` in combination with `summarize()` (figure \ref{fig:titv-viz}).

@@

We have shown how to create expressive and 
reproducible genomic workflows using the `plyranges` DSL. By
realising that the GRanges data model is tidy we have highlighted how to 
implement a grammar for performing genomic arithmetic, aggregation, restriction 
and merging. Our examples show that `plyranges` code is succinct, human
readable and can take advantage of the interoperability provided by
the Bioconductor ecosystem and the R language.

### 2018-05-16 paper.tex
- spelling fixes - making examples flow better - consistently use \texttt for package names

As with other genomic grammars, we define set operations that treat
ranges as sets of integers, including \emph{intersect}, \emph{union},
\emph{difference}, and \emph{complement}. There are two sets of these:
parallel and merging. The parallel operations map to infix operators,
which we surround with \emph{\%} symbols, a convention borrowed from R
syntax. For example, the parallel intersection (\emph{x \%intersect\%
y}) finds the intersecting range between \emph{xi} and \emph{yi} for
\emph{i} in \emph{1\ldots{}n}, where \emph{n} is the length of both
\emph{x} and \emph{y}. In constrast, the merging intersection
(\emph{intersect\_ranges(x, y)}) returns a new set of disjoint ranges
representing whereever there was overlap between a range in \emph{x} and
a range in \emph{y}. We use the infix syntax for the parallel
operations, since it is the conventional syntax for parallel, binary
operations in R. Finding the parallel union will fail when two ranges
have a gap, so we introduce a \emph{span} operator that takes the union
while filling any gap. The \emph{complement} operation is unique in that
it is unary. It finds the regions not covered by any of the ranges in a
single set. Closely related is the \emph{between} parallel operation,
which finds the gap separating \emph{xi} and \emph{yi}. The binary
operations are callable from within arithmetic, restriction and
aggregation expressions.

@@

The design of plyranges adheres to well understood principles of
language and API design: cognitive consistency, cohesion, expressiveness
and endomorphism \cite{Green1996-qg}. To varying degrees, these
principles also underly the design of dplyr and the Bioconductor
infrastructure.

@@

We aim for our interfaces to have a simple and direct mapping to the
user's cognitive model, i.e., how the user thinks about the data. This
requires careful selection of the level of abstraction so that the user
can express workflows in the language of genomics. This motivates the
adoption of the tidy GRanges object as our central data structure. The
basic data.frame and dplyr tibble lack any notion of genomic ranges and
so could not easily support our genomic grammar, with its specific verbs
for range-oriented data manipulation. Another example of cognitive
consistency is how plyranges is insensitive to direction/strand by
default when, e.g., detecting overlaps. GenomicRanges has the opposite
behavior. We believe that defaulting to purely spatial overlap is most
intuitive to most users.

Like dplyr, plyranges verbs are functional: they are free of side
effects and return their result. This enables chaining of verbs through
syntax like the forward pipe operator (\texttt{\%\textgreater{}\%}, read
as ``then'') of the magrittr package \cite{R-magrittr}. This has syntax
has a direct cognitive mapping to natural language and the intuitive
notion of pipelines. The low-level object-oriented APIs of Bioconductor
tend to manipulate data via subreplacement functions, like
\texttt{start(gr)\ \textless{}-\ x}. These ultimately produce the side
effect of replacing a symbol mapping in the current environment and thus
are not amenable to so-called fluent syntax.

@@

A function is cohesive if it performs a singular task without producing
any side-effects. Singular tasks are not necessarily atomic; they can
always be broken down further at lower levels of abstraction. For
example, to resize a range, the user needs to specify which position
(start, end, midpoint) should be invariant over the transformation. The
\texttt{resize()} function from the GenomicRanges package has a
\texttt{fix} argument that sets the anchor, so calling \texttt{resize()}
coalesces anchoring and width modification. The coupling at the function
call level is justified since the effect of setting the width depends on
the anchor. However, plyranges increases cohesion and decouples the
anchoring into its own function call.

@@

Expressiveness relates to the information content in code: the
programmer should be able to clarify intent without unnecessary
verbosity. For example, our overlap-based join operations are more
concise than the multiple steps necessary to achieve the same effect in
the original GenomicRanges API. In other cases, the plyranges API
increases verbosity for the sake of clarity and cohesion. Explicitly
calling \texttt{anchor()} can require more typing, but the code is
easier to comprehend. Another example is the set of routines for
importing genomic annotations, including
\texttt{read\textbackslash{}\_gff()},
\texttt{read\textbackslash{}\_bed()}, and \texttt{read\_bam()}. Compared
to the generic \texttt{import()} in rtracklayer, the explicit
format-based naming in plyranges clarifies intent and the type of data
being returned. Similarly, every plyranges function that computes with
strand information indicates its intentions by including suffixes such
as \texttt{directed}, \texttt{upstream} or \texttt{downstream} in its
name, otherwise strand is ignored. The GenomicRanges API does not make
this distinction explicit in its function naming, instead relying on a
parameter that defaults to strand sensitivity, an arguably unintuitive
behavior.

@@

The Bioconductor package \texttt{AnnotationHub} \cite{R-ahub} can be
used to search for BigWig files from ChIP-Seq experiments from the Human
Epigenome Roadmap project \cite{Roadmap-Epigenomics-Consortium2015-pr}.
Here we focus on assays for primary T CD8+ memory cells from peripheral
blood. Using \texttt{plyranges} we will read the BigWig file
corresponding to the H3 lysine 27 trimethylation (H3K27Me3) methylation
mark over chromosome 10.

@@

After filtering for regions with a coverage score greater than 8, we can
reduce individual runs to ranges representing the islands of coverage,
using the \texttt{reduce\_ranges()} function. This function allows a
summary to be computer over each island: in this case we take the
maximum of the scores to find the coverage peaks over chromosome 10.

@@

We can directly obtain the genome information from the header of the BAM
file: in this case the reads were aligned to the hg19 genome build and
there are no reads overlapping the mitochondrial genome. To generate
bins of fixed width 10000bp we apply \texttt{tile()} function from
\texttt{GenomicRanges} to the genomic coordinates (with \texttt{unlist}
to return a GRanges object)

@@

\NormalTok{bins <-}\StringTok{ }\NormalTok{locations }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{tile}\NormalTok{(}\DataTypeTok{width =}\NormalTok{ 10000L) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{unlist}\NormalTok{(}\DataTypeTok{use.names =} \OtherTok{FALSE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Next we only read in alginments that overlap the genomic locations we
are interested in and select the query seqeunce. Note that the reading
of the BAM file is deferred: only alignments that pass the filter are
loaded into memory. We can add another column representing the GC
proportion for each alignment using the \texttt{letterFrequency()}
function from the \texttt{Biostrings} package \cite{R-biostrings}.

@@

Finally, we use \texttt{group\_by\_overlaps()} with \texttt{summarize()}
to compute the total number of reads and averge GC content within each
genomic window.

@@

We can then compute the per-chromosome transition-transversion ratio
using the \texttt{group\_by} and \texttt{summarize} pattern (figure
\ref{fig:titv-viz}).

@@

We have shown how to create expressive and reproducible genomic
workflows using the \texttt{plyranges} DSL. By realising that the
GRanges data model is tidy we have highlighted how to implement a
grammar for performing genomic arithemtic, aggregation, restriction and
merging. Our examples show that \texttt{plyranges} code is succinct,
human readable and can take advantage of the interoperability provided
by the Bioconductor ecosytem and the R language.

>>>>

As with other genomic grammars, we define set operations that treat
ranges as sets of integers, including \emph{intersect}, \emph{union},
\emph{difference}, and \emph{complement}. There are two sets of these:
parallel and merging. The parallel operations map to infix operators,
which we surround with \emph{\%} symbols, a convention borrowed from R
syntax. For example, the parallel intersection (\emph{x \%intersect\%
y}) finds the intersecting range between \emph{xi} and \emph{yi} for
\emph{i} in \emph{1\ldots{}n}, where \emph{n} is the length of both
\emph{x} and \emph{y}. In contrast, the merging intersection
(\emph{intersect\_ranges(x, y)}) returns a new set of disjoint ranges
representing wherever there was overlap between a range in \emph{x} and
a range in \emph{y}. We use the infix syntax for the parallel
operations, since it is the conventional syntax for parallel, binary
operations in R. Finding the parallel union will fail when two ranges
have a gap, so we introduce a \emph{span} operator that takes the union
while filling any gap. The \emph{complement} operation is unique in that
it is unary. It finds the regions not covered by any of the ranges in a
single set. Closely related is the \emph{between} parallel operation,
which finds the gap separating \emph{xi} and \emph{yi}. The binary
operations are callable from within arithmetic, restriction and
aggregation expressions.

@@

The design of \texttt{plyranges} adheres to well understood principles
of language and API design: cognitive consistency, cohesion,
expressiveness and endomorphism \cite{Green1996-qg}. To varying degrees,
these principles also underlie the design of \texttt{dplyr} and the
Bioconductor infrastructure.

@@

We aim for our interfaces to have a simple and direct mapping to the
user's cognitive model, i.e., how the user thinks about the data. This
requires careful selection of the level of abstraction so that the user
can express workflows in the language of genomics. This motivates the
adoption of the tidy GRanges object as our central data structure. The
basic data.frame and \texttt{dplyr} tibble lack any notion of genomic
ranges and so could not easily support our genomic grammar, with its
specific verbs for range-oriented data manipulation. Another example of
cognitive consistency is how \texttt{plyranges} is insensitive to
direction/strand by default when, e.g., detecting overlaps.
\texttt{GenomicRanges} has the opposite behavior. We believe that
defaulting to purely spatial overlap is most intuitive to most users.

Like \texttt{dplyr}, \texttt{plyranges} verbs are functional: they are
free of side effects and return their result. This enables chaining of
verbs through syntax like the forward pipe operator
(\texttt{\%\textgreater{}\%}, read as ``then'') of the \texttt{magrittr}
package \cite{R-magrittr}. This has syntax has a direct cognitive
mapping to natural language and the intuitive notion of pipelines. The
low-level object-oriented APIs of Bioconductor tend to manipulate data
via sub-replacement functions, like \texttt{start(gr)\ \textless{}-\ x}.
These ultimately produce the side effect of replacing a symbol mapping
in the current environment and thus are not amenable to so-called fluent
syntax.

@@

A function is cohesive if it performs a singular task without producing
any side-effects. Singular tasks are not necessarily atomic; they can
always be broken down further at lower levels of abstraction. For
example, to resize a range, the user needs to specify which position
(start, end, midpoint) should be invariant over the transformation. The
\texttt{resize()} function from the \texttt{GenomicRanges} package has a
\texttt{fix} argument that sets the anchor, so calling \texttt{resize()}
coalesces anchoring and width modification. The coupling at the function
call level is justified since the effect of setting the width depends on
the anchor. However, plyranges increases cohesion and decouples the
anchoring into its own function call.

@@

Expressiveness relates to the information content in code: the
programmer should be able to clarify intent without unnecessary
verbosity. For example, our overlap-based join operations are more
concise than the multiple steps necessary to achieve the same effect in
the original \texttt{GenomicRanges} API. In other cases, the
\texttt{plyranges} API increases verbosity for the sake of clarity and
cohesion. Explicitly calling \texttt{anchor()} can require more typing,
but the code is easier to comprehend. Another example is the set of
routines for importing genomic annotations, including
\texttt{read\textbackslash{}\_gff()},
\texttt{read\textbackslash{}\_bed()}, and \texttt{read\_bam()}. Compared
to the generic \texttt{import()} in \texttt{rtracklayer}, the explicit
format-based naming in plyranges clarifies intent and the type of data
being returned. Similarly, every plyranges function that computes with
strand information indicates its intentions by including suffixes such
as \texttt{directed}, \texttt{upstream} or \texttt{downstream} in its
name, otherwise strand is ignored. The \texttt{GenomicRanges} API does
not make this distinction explicit in its function naming, instead
relying on a parameter that defaults to strand sensitivity, an arguably
unintuitive behavior.

@@

In the workflow of ChIP-seq data analysis, we are interested in finding
peaks from islands of coverage over chromosome. Here we will use
\texttt{plyranges} to call peaks from islands of coverage above 8 then
plot the region surrounding the tallest peak.

Using \texttt{plyranges} and the the Bioconductor package
\texttt{AnnotationHub} \cite{R-ahub} we can download and read BigWig
files from ChIP-Seq experiments from the Human Epigenome Roadmap project
\cite{Roadmap-Epigenomics-Consortium2015-pr}. Here we analyse a BigWig
file corresponding to H3 lysine 27 trimethylation (H3K27Me3) of primary
T CD8+ memory cells from peripheral blood, focussing on coverage islands
over chromosome 10.

@@

After filtering for regions with a coverage score greater than 8, we can
reduce individual runs to ranges representing the islands of coverage by
using the \texttt{reduce\_ranges()} function. This function allows a
summary to be computer over each island: in this case we take the
maximum of the scores to find the coverage peaks over chromosome 10.

@@

We can directly obtain the genome information from the header of the BAM
file: in this case the reads were aligned to the hg19 genome build and
there are no reads overlapping the mitochondrial genome. To generate
bins of fixed width 10000nt we apply the \texttt{tile\_ranges()}
function to the genomic coordinates in locations.

@@

\NormalTok{bins <-}\StringTok{ }\NormalTok{locations }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{tile_ranges}\NormalTok{(}\DataTypeTok{width =}\NormalTok{ 10000L)}
\end{Highlighting}
\end{Shaded}

Next we only read in alignments that overlap the genomic locations we
are interested in and select the query sequence. Note that the reading
of the BAM file is deferred: only alignments that pass the filter are
loaded into memory. We can add another column representing the GC
proportion for each alignment using the \texttt{letterFrequency()}
function from the \texttt{Biostrings} package \cite{R-biostrings}.

@@

Finally, we use \texttt{group\_by\_overlaps()} to see where the
alignments overlap the genomic windows and then apply
\texttt{summarize()} to compute the total number of reads and average GC
content within each window.

@@

We then compute the transition-transversion ratio over each chromosome
using \texttt{group\_by()} in combination with \texttt{summarize()}
(figure \ref{fig:titv-viz}).

@@

We have shown how to create expressive and reproducible genomic
workflows using the \texttt{plyranges} DSL. By realising that the
GRanges data model is tidy we have highlighted how to implement a
grammar for performing genomic arithmetic, aggregation, restriction and
merging. Our examples show that \texttt{plyranges} code is succinct,
human readable and can take advantage of the interoperability provided
by the Bioconductor ecosystem and the R language.
